---
title: "PA1_template"
author: "Javier Rios Perez"
date: "1/14/2017"
output: html_document
---
This is the solution for the Reproducible research - week 2
course.

Loading the necesary libraries
```{r, echo = TRUE}
library(ISLR)
library(plyr)
library(dplyr)
library(pander)
library(ggplot2)
```

Loading and processing the data

```{r}
original_data <- read.csv("activity.csv", header = TRUE)
monitoring_data <- original_data
```

Converting the data column into a date variable

```{r, echo = TRUE}
monitoring_data$date <- as.character(monitoring_data$date)
monitoring_data$date <- as.Date(monitoring_data$date, format = "%Y-%m-%d")
```

# What is mean total number of steps taken per day?
1. Calculate the total number of steps taken per day

```{r, echo = TRUE}

total_num_steps <- as.data.frame(ddply(monitoring_data,~date,summarise,
                                       sum=sum(steps, na.rm = TRUE)))
head(total_num_steps)
```
2. If you do not understand the difference between a histogram and a barplot,
research the difference between them. Make a histogram of the total number of 
steps taken each day

```{r, echo = TRUE}

png("plot1.png",
    width = 480,
    height = 480,
    units = "px")

hist(total_num_steps$sum, breaks = 20, col = "green", main = "Total number of steps",
     xlab = "Total number of steps")

dev.off()
```

3. Calculate and report the mean and median of the total
#number of steps taken per day

```{r, echo = TRUE}
org_steps_per_day <- ddply(monitoring_data,~date,summarise,org_sum=sum(steps))
mean_median_original <- ddply(monitoring_data,~date,summarise,mean=mean(steps),
                              median=median(steps))

head(mean_median_original)
tail(mean_median_original)
```
# What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of
the 5-minute interval (x-axis) and the average number of
steps taken, averaged across all days (y-axis)

```{r, echo = TRUE}
org_num_steps_intervals <- data.frame()
org_num_intervals <- length(unique(monitoring_data$interval))

for(i in 1:org_num_intervals)
{
  org_steps_interval <- filter(monitoring_data, interval == monitoring_data$interval[i])
  org_num_steps_intervals[i,1] <- mean(org_steps_interval$steps, na.rm = TRUE)
  org_num_steps_intervals[i,2] <- org_steps_interval$interval[1]
}


png("plot2.png",
    width = 480,
    height = 480,
    units = "px")

plot(org_num_steps_intervals[,1], type = "l", col = "red", xlab = "5-minute interval",
     ylab = "Averaged number of steps averaged accross all days",
     main = "5-minute interval Vs. Averaged number of steps")

dev.off()

```

2. Which 5-minute interval, on average across
all the days in the dataset, contains the maximum number of steps?

```{r, echo = TRUE}
org_max <- max(org_num_steps_intervals[,1])
org_max_interval_aux <- which(org_num_steps_intervals[,1] == org_max)
org_num_steps_intervals[org_max_interval_aux,2]
```

# Imputing missing values

1. Calculate and report the total number of missing values in the dataset
(i.e. the total number of rows with NAs)

```{r, echo = TRUE}
total_NA <- is.na(monitoring_data$steps)
table(total_NA)

png("plot3.png",
    width = 480,
    height = 480,
    units = "px")

barplot(table(total_NA), main = "Total number of missing (NA) step values")

dev.off()

```

2. Devise a strategy for filling in all of the missing values in the dataset.
The strategy does not need to be sophisticated. For example, you could use the
mean/median for that day, or the mean for that 5-minute interval, etc.

```{r, echo = TRUE}
aux <- unique(monitoring_data$date)
monitoring_clean <- data.frame()

for(i in 1: length(aux))
  
{
  day <- monitoring_data[monitoring_data$date == unique(monitoring_data$date)[i],]
  if(any(is.na(day)))
  {
    temp <- mean(day$steps, na.rm = TRUE)
    if(is.nan(temp))
    {
      temp <- 0
    }
    
  }
  # here i need to replace the NA <- temp for each day, remember to change and seve
  # the results in a data frame
  for(j in 1:dim(day)[1])
  {
    if(is.na(day[j,1]))
    {
      day[j,1] <- temp
    }
  }
  monitoring_clean <- rbind(monitoring_clean,day)
}

head(monitoring_clean)
```
3. Create a new dataset that is equal to the original
dataset but with the missing data filled in.

```{r, echo = TRUE}
head(monitoring_clean)
tail(monitoring_clean)
```

4. Make a histogram of the total number of steps taken
each day and Calculate and report the mean and median total
number of steps taken per day. Do these values differ from the
estimates from the first part of the assignment? What is the impact
of imputing missing data on the estimates of the total daily number of steps?
What is mean total number of steps taken per day

```{r, echo = TRUE}
steps_per_day <- ddply(monitoring_clean,~date,summarise,sum=sum(steps))

png("plot4.png",
    width = 480,
    height = 480,
    units = "px")

plot(steps_per_day, main = "Total number of steps per day", col = "blue")

dev.off()

# plotting the histogram

png("plot5.png",
    width = 480,
    height = 480,
    units = "px")

hist(steps_per_day$sum, breaks = 20, col = "green4", main = "Total number of steps",
     xlab = "Total number of steps")

dev.off()

# plotting the histogram

png("plot6.png",
    width = 480,
    height = 480,
    units = "px")

par(mfrow = c(2,1))
hist(total_num_steps$sum, breaks = 20, col = "green", main = "Total number of steps - Ignoring NA values",
     xlab = "Total number of steps")
hist(steps_per_day$sum, breaks = 20, col = "green4", main = "Total number of steps - with replacement",
     xlab = "Total number of steps")

dev.off()

# Calculate and report the mean and median of the total number of steps taken per day

mean_median_with_substitution <- ddply(monitoring_clean,
                                       ~date,summarise,mean_sub=mean(steps),median_sub=median(steps))

# making a new data frame that does not contain missing values
mean_median_combined <- cbind(mean_median_original,
                              mean_median_with_substitution$mean_sub, mean_median_with_substitution$median_sub)

# comparing the total number of steps

total_number_steps_combined <- cbind(org_steps_per_day, steps_per_day[,2])

non_zeros <- monitoring_clean[-which(monitoring_clean$steps == 0),]
daily_non_zeros <- ddply(non_zeros,~date,summarise,sum=sum(steps))

png("plot7.png",
    width = 480,
    height = 480,
    units = "px")

plot(daily_non_zeros, main = "Total number of steps per day")

dev.off()

```
# Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend”
indicating whether a given date is a weekday or weekend day.

```{r, echo = TRUE}
weekdays <- data.frame() # i am going to store the resutls of weekend/weekend analysis

for(i in 1:length(monitoring_clean$date))
  
{
  if(weekdays(monitoring_clean[i,2]) == "Saturday" | weekdays(monitoring_clean[i,2]) == "Sunday")
  {
    temp <- "weekend"
  }
  
  else
  {
    temp <- "weekday"
  }
  
  weekdays[i,1] <- temp
  i <- i+1
}

weekly_steps_per_day <- as.data.frame(cbind(monitoring_clean, weekdays))
colnames(weekly_steps_per_day)[4] <- "weekdays"
weekly_steps_per_day$weekdays <- as.factor(weekly_steps_per_day$weekday)


num_steps_intervals_weekdays <- data.frame()
avg_steps_intervals_weekdays <- data.frame()

num_steps_intervals_weekends <- data.frame()
avg_steps_intervals_weekends <- data.frame()

num_intervals_weekdays <- length(unique(weekly_steps_per_day$interval))

weekdays_data <- filter(weekly_steps_per_day, weekly_steps_per_day$weekdays == "weekday")
weekends_data <-  filter(weekly_steps_per_day, weekly_steps_per_day$weekdays == "weekend")

for(i in 1:num_intervals_weekdays)
{
  num_steps_intervals_weekdays <- filter(weekdays_data, interval == weekdays_data$interval[i])
  avg_steps_intervals_weekdays[i,1] <- mean(num_steps_intervals_weekdays$steps)
  avg_steps_intervals_weekdays[i,2] <- num_steps_intervals_weekdays$interval[2]
  avg_steps_intervals_weekdays[i,3] <- num_steps_intervals_weekdays$weekdays[4]
}

for(i in 1:num_intervals_weekdays)
{
  num_steps_intervals_weekends <- filter(weekends_data, interval == weekends_data$interval[i])
  avg_steps_intervals_weekends[i,1] <- mean(num_steps_intervals_weekends$steps)
  avg_steps_intervals_weekends[i,2] <- num_steps_intervals_weekends$interval[2]
  avg_steps_intervals_weekends[i,3] <- num_steps_intervals_weekends$weekdays[4]
}

# changing col names for both weekdays and weekend new dataframes

colnames(avg_steps_intervals_weekdays) <- c("avg", "intervals", "weekday")
colnames(avg_steps_intervals_weekends) <- c("avg", "intervals", "weekday")

# plotting the results

png("plot8.png",
    width = 480,
    height = 480,
    units = "px")

par(mfrow = c(2,1))

plot(avg_steps_intervals_weekdays$avg, type = "l", col = "red", xlab = "Intervals", ylab = "Avg. Steps",
     main = " Weekdays average steps per interval of time")
plot(avg_steps_intervals_weekends$avg, type = "l", col = "blue", xlab = "Intervals", ylab = "Avg. Steps",
     main = " Weekends average steps per interval of time")

dev.off()
```


